---
title: "Customer Segmentation for UK retailer"
authors: "Sanjana Chalavadi"
date: '2022-12-07'
---


```{r, loading_required_libraries}
#loading all the required libraries
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
library(dplyr)
library(magrittr)
library(ggplot2)
library(plotly)
library(data.table)
library(tidyr)
library(cowplot)
library(cluster)
library(GGally)
library(tidyverse)
library(readxl)
library(DataExplorer)
library(visdat)
library(naniar)
library(rmarkdown)
library(knitr) 
library(DT)
library(wordcloud2) 
library(wordcloud) 
library(extrafont) 
library(SnowballC) 
library(tm) 
library(dlookr) 
library(gridExtra)
library(scales)
library(lubridate) 
library(ggThemeAssist)
library(hrbrthemes)
library(xts)
library(highcharter)
library(countrycode)
library(factoextra)
library(dplyr)
library(funModeling)
```


```{r,loading_data}
#reading in the dataframe and checking the answers
df <- read.csv("Online_Retail_v2.csv", header=TRUE)
head(df)
```

```{r,structure_of_data}
str(df)
summary(df)
head(df, 5)
n_distinct(df$CustomerID)
n_distinct(df$Description)
```

```{r, missing_values_check}
gg_miss_var(df, show_pct = TRUE)
```
```{r, removing_missing_values}
#Removing missing values in customer_id
df <- df %>% na.omit(df$CustomerID)

# Replacing NA Description values with the string "empty"
df$Description <- replace_na(df$Description, "No-info")
```


```{r, outlier_treatment}
plot_outlier(df, Quantity, UnitPrice, col = "#ACAC89")
min(df$UnitPrice)
max(df$UnitPrice)
```

```{r,feature_extraction_spent}
#Feature extraction
#Step 1: spent column -> total quantity * unit price
df_fe <- df
df_fe <- mutate(df_fe,Spent=Quantity*UnitPrice)
head(df_fe)

```

```{r,feature_extraction_date_time}
#step 2: we are extracting derived columns like weekday, hr, month,year from the date column
#converting the character invoice date to date format
df_fe$InvoiceDate <- dmy_hm(df_fe$InvoiceDate)

#date extraction
df_fe$year <- year(df_fe$InvoiceDate)
df_fe$month <- month(df_fe$InvoiceDate)
df_fe$hour <- hour(df_fe$InvoiceDate)
df_fe$wday <- wday(df_fe$InvoiceDate,label = TRUE)
```


```{r,eda1}
#Bar plot - Country wise revenue earned
p1 <- df_fe %>% group_by(Country) %>% dplyr::summarise(total_revenue = sum(Spent))
a1 <- ggplot(data = p1, mapping = aes(x = reorder(Country, -total_revenue), total_revenue)) + 
  geom_bar(stat = "identity",fill="dodgerblue2") + coord_flip()+  labs( x = 'Revenue earned',y = "Country", title = "Total Revenue by Country") +theme(axis.text.x = element_text(angle=90))

p2 <- df_fe %>% group_by(Country) %>% filter(Country != "United Kingdom") %>% dplyr::summarise(total_revenue = sum(Spent))
a2 <- ggplot(data = p2, mapping = aes(x = reorder(Country, -total_revenue), total_revenue)) + 
  geom_bar(stat = "identity",fill="dodgerblue2") +  coord_flip()+ labs( x = 'Revenue',y = "Country", title = "(Excluding UK)")+theme(axis.text.x = element_text(angle=90))

plot_grid( a1,a2, ncol = 2) + labs (title = "Total Revenue by Country") +theme(plot.title = element_text(hjust = 0.5,face = "bold"))
```

```{r,eda2}
#Barplot - Revenue count by Weekday
p3 <- df_fe %>% group_by(wday) %>% dplyr::summarise(total_revenue = sum(Spent))
a3 <- ggplot(data = p3, mapping = aes(x = reorder(wday, -total_revenue), total_revenue)) + 
  geom_bar(stat = "identity",fill="coral4") +   labs( y = 'Revenue',x = "Weekday", title = "Total Revenue by Weekday") +theme(axis.text.x = element_text(angle=90))

p4 <- df_fe %>% group_by(hour) %>% dplyr::summarise(total_revenue = sum(Spent))
a4 <- ggplot(data = p4, mapping = aes(x = reorder(hour, -total_revenue), total_revenue)) + 
  geom_bar(stat = "identity",fill="orchid4") +  labs( y = 'Revenue',x = "Hour", title = "Total Revenue by Hour") +theme(axis.text.x = element_text(angle=90))

p5 <- df_fe %>% group_by(month) %>% dplyr::summarise(total_revenue = sum(Spent))
a5 <- ggplot(data = p5, mapping = aes(x = reorder(month, -total_revenue), total_revenue)) + 
  geom_bar(stat = "identity",fill="seagreen4") +   labs( y = 'Revenue',x = "Month", title = "Total Revenue by Month") +theme(axis.text.x = element_text(angle=90))

p6 <- df_fe %>% group_by(year) %>% dplyr::summarise(total_revenue = sum(Spent))
a6 <- ggplot(data = p6, mapping = aes(x = reorder(year, -total_revenue), total_revenue)) + 
  geom_bar(stat = "identity",fill="orangered2") +   labs( y = 'Revenue',x = "Year", title = "Total Revenue by Year") +theme(axis.text.x = element_text(angle=90))


plot_grid( a3,a4,a5,a6, ncol = 2,nrow=2) + labs (title = "Total Revenue by Time") +theme(plot.title = element_text(hjust = 0.5,face = "bold"))
```

```{r,customers}
p7 <- df_fe %>% group_by(CustomerID) %>% dplyr::summarise(amount_spent = sum(Spent)) %>% arrange(-amount_spent)
ggplot(data = head(p7,10), mapping = aes(x = reorder(as.factor(CustomerID), -amount_spent), amount_spent)) + 
  geom_bar(stat = "identity",fill="turquoise4")+  labs( y = 'Revenue',x = "Customer ID", title = "Top 10 Customers") +theme(axis.text.x = element_text(angle=90))


p8 <- df_fe %>% group_by(Description) %>% dplyr::summarise(amount_spent = sum(Spent)) %>% arrange(-amount_spent)
ggplot(data = head(p8,10), mapping = aes(x = reorder((Description), -amount_spent), amount_spent)) + 
  geom_bar(stat = "identity",fill="orange1")+  labs( y = 'Revenue',x = "Products", title = "Top 10 Products") +theme(axis.text.x = element_text(angle=90)) + coord_flip()

```

```{r,cancel_order_analysis}
#cancelled orders by country, weekday, hour, products
cancel_df <- df_fe %>% filter(Quantity < 0)

p9 <- cancel_df %>% group_by(CustomerID) %>% dplyr::summarise(amount_spent = sum(Spent)) %>% arrange(amount_spent)
ggplot(data = head(p9,10), mapping = aes(x = reorder(as.factor(CustomerID), -amount_spent), amount_spent)) + 
  geom_bar(stat = "identity",fill="violetred4")+  labs( y = 'Revenue',x = "Customer ID", title = "Top 10 Customer's cancelling Orders") +theme(axis.text.x = element_text(angle=90))


p10 <- cancel_df %>% group_by(Description) %>% dplyr::summarise(amount_spent = sum(Spent)) %>% arrange(amount_spent)
ggplot(data = head(p10,10), mapping = aes(x = reorder((Description), -amount_spent), amount_spent)) + 
  geom_bar(stat = "identity",fill="orange4")+  labs( y = 'Revenue',x = "Products", title = "Top 10 Products being cancelled") +theme(axis.text.x = element_text(angle=90)) + coord_flip()
```

```{r,word_cloud_EDA}
# preparing and cleaning the text
products_list <- unique(df_fe$Description)
docs <- Corpus(VectorSource(products_list))
toSpace <- content_transformer(function (x , pattern) gsub(pattern, " ", x))
docs <- tm_map(docs, toSpace, "/")
docs <- tm_map(docs, toSpace, "@")
docs <- tm_map(docs, toSpace, "\\|")
# Convert the text to lower case
docs <- tm_map(docs, content_transformer(tolower))
# Remove numbers
docs <- tm_map(docs, removeNumbers)
# Remove english common stopwords
docs <- tm_map(docs, removeWords, stopwords("english"))
# Removing own stop words
docs <- tm_map(docs, removeWords, c("pink", "blue","red","set","white","metal", "glass","large","small","holder","black","design"))
# Remove punctuations
docs <- tm_map(docs, removePunctuation)
# Eliminate extra white spaces
docs <- tm_map(docs, stripWhitespace)

dtm <- TermDocumentMatrix(docs)
m <- as.matrix(dtm)
v <- sort(rowSums(m),decreasing=TRUE)
d <- data.frame(word = names(v),freq=v)

set.seed(123)
wordcloud(words = d$word, freq = d$freq, min.freq = 1,
          max.words=20, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))

```

```{r,revenue_map}
map_info <- df_fe %>% group_by(Country) %>%  dplyr::summarise(revenue = sum(Spent))
highchart(type = "map") %>%
  hc_add_series_map(worldgeojson,
                    map_info %>% 
                      bind_cols(as_tibble(map_info$revenue)) %>% 
                      group_by(map_info$Country) %>% 
                      dplyr::summarise(revenue = log1p(sum(value))) %>% 
                      ungroup() %>% 
                      mutate(iso2 = countrycode(sourcevar = map_info$Country, 
                          origin="country.name", destination="iso2c")),
                    value = "revenue", joinBy = "iso2") %>%
  hc_title(text = "Revenue by country (log)") %>%
  hc_tooltip(useHTML = TRUE, headerFormat = "",
             pointFormat = "{point.map_info$Country}") %>% 
  hc_colorAxis(stops = color_stops(colors = viridisLite::turbo(10, begin = 0.1)))
```


# MODELING 

```{r,RFM}
## recency
recency <- df_fe %>% dplyr::select(CustomerID, InvoiceDate) %>% mutate(recency = as.Date("2011-12-09") - as.Date(InvoiceDate))  
recency <- recency %>% dplyr::select(CustomerID, recency) %>% group_by(CustomerID) %>% slice(which.min(recency))

#frequency
amount_products <- df_fe %>%dplyr::select(CustomerID, InvoiceDate) %>% group_by(CustomerID, InvoiceDate) %>% dplyr::summarise(n_prod = n())
df_frequency <- amount_products %>% dplyr::select(CustomerID) %>%group_by(CustomerID) %>% dplyr::summarise(frequency = n())

#monetary
customer <- summarise_at(group_by(df_fe,CustomerID,Country), vars(Spent,Quantity), funs(sum(.,na.rm = TRUE)))
monetary <- select(customer, c("CustomerID", "Spent"))

#RFM DF
# inner join the three RFM data frames by CustomerID
rfm <- recency %>% dplyr::inner_join(., df_frequency, by = "CustomerID") %>% dplyr::inner_join(., monetary, by = "CustomerID")
# drop the days from recency column and transform it into numeric data type
rfm <- rfm %>% mutate(recency = str_replace(recency, " days", "")) %>% mutate(recency = as.numeric(recency)) %>% ungroup()
head(rfm, 3)
```


```{r,rfm_plot}
h1 <- hchart(rfm$recency, color = "#4286f4", name = "Recency")
h2 <- hchart(rfm$frequency, color = "#d66048", name = "Frequency")
h3 <- hchart(rfm$Spent, color = "#4A6670", name = "Monetary Value")
htmltools::browsable(hw_grid(h1, h2, h3, ncol = 3, rowheight = 500))
```


```{r, K_means_clustering}
rfm1 <- select(rfm, -CustomerID)
df_scale <- scale(rfm1)

#elbow method
set.seed(123) 
km_out_list <- lapply(1:20, function(k) list(k=k,km_out=kmeans(df_scale, k, nstart = 30)))
km_results <- data.frame(
  k=sapply(km_out_list, function(k) k$k),
  totss=sapply(km_out_list, function(k) k$km_out$totss),
  tot_withinss=sapply(km_out_list, function(k) k$km_out$tot.withinss)
  )

plot_ly(km_results,x=~k,y=~tot_withinss,mode = 'lines+markers')
```

Inference: 
Typically, the elbow point or change in slope from steep to gentle is where we consider k to be at its best. Such points are observed here with k = 3, 5 and 7.Despite the fact that the graph suggests 3 over 5 & 7(based on slope change from steep to gentle), it may not be the ideal value because it overly generalizes the customers and prevents us from performing targeted marketing. Therefore, we can eliminate considering 3 as no. of clusters after taking domain knowledge into account.

```{r,kmeans_silhoutte}
#Silhouette
set.seed(123)
fviz_nbclust(df_scale,kmeans, method = "silhouette")+labs(subtitle = "Silhouette method") 
km.res <- kmeans(df_scale, 10, nstart = 20)
```

```{r,kmeans_gap_statistic}
library(cluster)
# compute gap statistic
set.seed(123)
gap_stat <- suppressWarnings(clusGap(df_scale, FUN = kmeans, nstart = 25,K.max = 10, B = 50))
fviz_gap_stat(gap_stat) + labs(subtitle = "Gap statistic method")
```
Inference: 
- Elbow plot suggest 3,5 & 7.
- Silhouette plot suggests 4. 
- Gap-statistic is not converging. But in the selected range local maxima was attained at 5. 

We eliminated 3 as our objective is targeted advertising. Considering 5 seems an appropriate through elbow and gap statistics.


```{r,kmeans_cluster_char}
set.seed(123)
km.res_5 <- kmeans(df_scale, 5, nstart = 10)
#Withiness score
km.res_5$betweenss/km.res_5$totss

rfm <- rfm %>% mutate(kmm_cluster = km.res_5$cluster)

#cluster analysis
fviz_cluster(km.res_5, geom = "point", data = df_scale) + ggtitle("Kmeans Clustering for Optimal K (k=5)")
hist(rfm$kmm_cluster,xlab = "Cluster",col = "blue",main= "Kmeans Cluster distribution")

#cluster characteristics
a <- ggplot(rfm, aes(x = kmm_cluster, y = log(frequency), group = kmm_cluster, fill = as.factor(kmm_cluster))) + 
  geom_boxplot(show.legend = FALSE) + theme_minimal() + scale_fill_brewer(palette = "Set2") 
b <- ggplot(rfm, aes(x = kmm_cluster, y = log(Spent), group = kmm_cluster, fill = as.factor(kmm_cluster))) + 
  geom_boxplot(show.legend = FALSE) + theme_minimal() + scale_fill_brewer(palette = "Set2")
c <- ggplot(rfm, aes(x = kmm_cluster, y = recency, group = kmm_cluster, fill = as.factor(kmm_cluster))) + 
  geom_boxplot(show.legend = FALSE) + theme_minimal() + scale_fill_brewer(palette = "Set2")
plot_grid( a,b,c, ncol = 3) + labs (title = "RFM Cluster Characteristics for KMeans") +theme(plot.title = element_text(hjust = 0.5,face = "bold"))
```

```{r,DB_Scan}
# Density Based cluster - To avoid the impact of outliers
#install.packages("fpc")
#install.packages("dbscan")
library("fpc")
library("factoextra")
set.seed(123)
db <- fpc::dbscan(df_scale, eps=0.15, MinPts = 5, scale = TRUE)
fviz_cluster(db, df_scale, stand = FALSE, ellipse = FALSE, show.clust.cent = FALSE,geom = "point",palette = "jco",ggtheme=theme_classic())
```
```{r,cluster_characterstics}

rfm <- rfm %>% mutate(db_cluster = db$cluster)

hist(rfm$db_cluster,xlab = "Cluster",col = "blue",main= "DBScan Cluster distribution")

#cluster characteristics
d <- ggplot(rfm, aes(x = db_cluster, y = log(frequency), group = db_cluster, fill = as.factor(db_cluster))) + 
  geom_boxplot(show.legend = FALSE) + theme_minimal() + scale_fill_brewer(palette = "Set2") 
e <- ggplot(rfm, aes(x = db_cluster, y = log(Spent), group = db_cluster, fill = as.factor(db_cluster))) + 
  geom_boxplot(show.legend = FALSE) + theme_minimal() + scale_fill_brewer(palette = "Set2")
f <- ggplot(rfm, aes(x = db_cluster, y = recency, group = db_cluster, fill = as.factor(db_cluster))) + 
  geom_boxplot(show.legend = FALSE) + theme_minimal() + scale_fill_brewer(palette = "Set2")
plot_grid( d,e,f, ncol = 3) + labs (title = "RFM Cluster Characteristics for DB Scan") +theme(plot.title = element_text(hjust = 0.5,face = "bold"))

```

```{r}

#Min distance clustering
kNNdist <- sort(dbscan::kNNdistplot(scale(df_scale), k =  5))
abline(h = 0.15, lty = 2)
```

```{r,gmm_clustering}
library(ClusterR)
set.seed(123)
opt_gmm = Optimal_Clusters_GMM(df_scale, max_clusters = 9, criterion = "BIC", dist_mode = "maha_dist", seed_mode = "random_subset",
                               km_iter = 10, em_iter = 10, var_floor = 1e-10,plot_data = T)

gmm = GMM(df_scale, 5, dist_mode = "maha_dist", seed_mode = "random_subset", km_iter = 10,em_iter = 10, verbose = F)  
pr = predict(gmm, newdata = df_scale)
```

```{r}
pr
```


```{r}
rfm <- rfm %>% mutate(gmm_cluster = pr)

hist(rfm$gmm_cluster,xlab = "Cluster",col = "blue",main= "GMM Cluster distribution")

#cluster characteristics
g <- ggplot(rfm, aes(x = gmm_cluster, y = log(frequency), group = gmm_cluster, fill = as.factor(gmm_cluster))) + 
  geom_boxplot(show.legend = FALSE) + theme_minimal() + scale_fill_brewer(palette = "Set2") 
h <- ggplot(rfm, aes(x = gmm_cluster, y = log(Spent), group = gmm_cluster, fill = as.factor(gmm_cluster))) + 
  geom_boxplot(show.legend = FALSE) + theme_minimal() + scale_fill_brewer(palette = "Set2")
i <- ggplot(rfm, aes(x = gmm_cluster, y = recency, group = gmm_cluster, fill = as.factor(gmm_cluster))) + 
  geom_boxplot(show.legend = FALSE) + theme_minimal() + scale_fill_brewer(palette = "Set2")
plot_grid( g,h,i, ncol = 3) + labs (title = "RFM Cluster Characteristics for GMM Clustering") +theme(plot.title = element_text(hjust = 0.5,face = "bold"))

```

```{r,RFM_Analysis}
recency_s <- df_fe %>% 
  dplyr::select(CustomerID, InvoiceDate) %>% 
  group_by(CustomerID) %>% 
  slice(which.max(InvoiceDate))

rfm_test <- customer %>% inner_join(recency_s, by = "CustomerID")

rfm_test <- rfm_test %>% left_join(rfm%>%select(CustomerID,recency,frequency,Spent), by = "CustomerID") 

library(rfm)
rfm_result <- rfm_table_customer(rfm_test, CustomerID, Quantity, recency, Spent.x, as.Date("2011-12-09"))
rfm_datatable <- rfm_result$rfm
```


```{r}
rfm_segments <- rfm_datatable
rfm_segments <- rfm_segments %>% 
  mutate(segment = ifelse(recency_score >= 4 & frequency_score >= 4 & monetary_score >= 4, "Champion", 
  ifelse(recency_score >= 2 & frequency_score >= 3 & monetary_score >= 3, "Loyal Customer", 
  ifelse(recency_score >= 3 & frequency_score <= 3 & monetary_score <= 3, "Potential Loyalist",
  ifelse(recency_score >= 4 & frequency_score <= 1 & monetary_score <= 1, "New Customer",
  ifelse((recency_score == 3 | recency_score == 4) & frequency_score <= 1 & monetary_score <= 1, "Promising",
  ifelse((recency_score == 2 | recency_score == 3) & (frequency_score == 2 | frequency_score == 3) & 
           (monetary_score == 2 | monetary_score == 3), "Need Attention",
  ifelse((recency_score == 2 | recency_score == 3) & frequency_score <= 2 & monetary_score <= 2, "About to Sleep",
  ifelse(recency_score <= 2 & frequency_score > 2 & monetary_score > 2, "At Risk",
  ifelse(recency_score <= 1 & frequency_score >= 4 & monetary_score >= 4, "Can't lose them",
  ifelse(recency_score <= 2 & frequency_score == 2 & monetary_score == 2, "Hibernating", "Lost")))))))))))
```

```{r}
ggplot(rfm_segments) + geom_bar(aes(x = segment, fill = segment))+theme(axis.text.x=element_text(angle=90,hjust=1)) +labs(title = "Barplot for Segments of customers")
```
```{r}
rfm_segments %>% mutate(segment = factor(segment, labels = c("Lost", " Champion", "Loyal Customer", "Hibernating", "Potential Loyalist","Need Attention","At Risk","About to Sleep")))

n <- ggplot(rfm_segments, aes(x = segment, y = recency_days))+ geom_boxplot()+theme(axis.text.x = element_text(angle=90))
o <- ggplot(rfm_segments, aes(x = segment, y = log(transaction_count)))+ geom_boxplot()+theme(axis.text.x = element_text(angle=90))
p <- ggplot(rfm_segments, aes(x = segment, y = log(amount)))+ geom_boxplot()+theme(axis.text.x = element_text(angle=90))

plot_grid(n,o,p, ncol = 3) + labs (title = "RFM Analysis Cluster Characteristics") +theme(plot.title = element_text(hjust = 0.5,face = "bold"))
```


```{r}
ggplotly(suppressWarnings(ggplot(rfm_segments, aes(x=as.factor(segment), y=recency_days, fill=as.factor(segment))))) + 
labs(title="Income distribution for each cluster",x="cluster", y = "Income") + theme_classic()+ geom_boxplot( outlier.shape=NA)
```


```{r,Hierarcial_Clustering}

fviz_nbclust(df_scale, hcut, method = "wss") +
geom_vline(xintercept = 3, linetype = 2)+
labs(subtitle = "Elbow method")
```

```{r}
fviz_nbclust(df_scale, hcut, method = "silhouette")+
labs(subtitle = "Silhouette method")
```

```{r}
# Gap statistic
# nboot = 50 to keep the function speedy.
# recommended value: nboot= 500 for your analysis.
# Use verbose = FALSE to hide computing progression.
set.seed(123)
fviz_nbclust(df_scale, hcut, nstart = 25, method = "gap_stat",
nboot = 50)+
labs(subtitle = "Gap statistic method")
```

```{r}
nb.complete <- NbClust(df_scale, distance = "manhattan", min.nc = 2,max.nc = 10, method = "complete")
# visualising the result
suppressWarnings( fviz_nbclust(nb.complete))
```


